### 2023-04-16 tex/article.tex
Some fix

Our approach combines daily satellite SST data, which are used to detect fronts and sort them by their strength, with daily satellite surface \as{chl}, from which we derive anomalies over fronts.
For \as{chl}, we used the \abbrv{L3}~product distributed by ACRI"~ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (\as{seawifs}, \as{modis} Aqua \& Terra, \as{meris}, \as{viirs}-\as{snpp} \& \as{jpss}[\textlf{-1}], \as{olci}[\abbrv{-S3A}] \& \abbrv{S3B}) merged and reprocessed, available daily at \qty{4}{\km} resolution (\cite{garnesson_2019})\footnote{Dataset \dataname{chl_globcolour} décrit~\nref[section]{sec:donnees-chl_globcolour}}.

For SST, we used the \al{esa} \al{sst} \al{cci} analysis product version 2.1 (\cite{merchant_2019, good_2020})\footnote{Dataset \dataname{sst_esacci} décrit \nref[section]{sec:donnees-sst_esacci}}, also available daily at \qty{4}{\km} resolution over the period 2000--2020.
This product combines data from all available infrared sensors (\as{atsr}, \as{slstr}, and \as{avhrr}), ensuring good resolution where data are available, unlike other SST products which also include microwave and in-situ measurements, resulting in considerable smoothing of the SST field. %chktex 36
Where SST data is not available, spatial interpolation is performed to obtain a cloud free product which, at the cost of resolution on finer features, provides complete synoptic coverage of our large study area.
This interpolation tends to underestimate the detection of fronts, as the SST field is smoother over cloud-covered areas (\cite{merchant_2019}).
However, the combination of several sensors allows to reduce these areas to a minimum.
Furthermore, we have only considered cloud-free pixels for our analysis, which ensures that cloudy areas are not taken into account in our quantification.

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig01.pdf}
  \caption[Delimitation of biomes]{
    Delimitation of the three biomes in the open ocean Gulf Stream extension region: the Permanent Subtropical Biome (PSB, south of the dashed line at \latlon{32N}), the Seasonal Subtropical Biome (SSB, between \latlon{32N} and the meandering Gulf Stream northern wall on that day marked with the black contour), and the sub-Polar Biome (PB, north of the Gulf Stream northern wall).
    (a)~SST and (b)~\as{chl} snapshots on the 22 April 2007 (with data masked by clouds in white). The red line follows the \qty{1500}{\m} isobath.
    Data on the continental shelf (\qty{<1500}{\m}) is not considered here and have been masked.
    (c)~SST and (d)~\as{chl} distribution within each biome for the same day (PB:~blue, SSB:~yellow, PSB:~red).
    The black line in (c) shows the SST threshold value detected to delimit the Gulf Stream northern wall (see methods section).
  }%
  \label{fig:zone-separation}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig02.pdf}
  \caption[Example of front]{
    The SST, \as{chl}, and heterogeneity index (HI) of a front on the 7 July 2007.
    The plain and dashed contours correspond to HI values of 5 and 10.
    This front is categorized as weak.
    \as{chl} are elevated inside the front.
  }%
  \label{fig:zoom}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig03.pdf}
  \caption[Distribution of Chl-\textit{a} against HI]{
    Normalized distribution of the Heterogeneity Index (HI, red dashed line) within each biome, and distribution of \as{chl} as a function of HI (representing front strength), over the full period 2000--2020.
    Shown are the median value of the \as{chl} distributions (solid black line), and 1st and 3rd quartiles (dashed lines).
    Note that \pct{0.5} of pixels have outstanding large HI values and are not included here.
  }%
  \label{fig:chl-vs-hi}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig04.pdf}
  \caption[Occurrence map of strong and weak fronts]{
    Occurrence of (a) weak fronts and (b) strong fronts expressed as the percentage of time steps over the time series (2000--2020) for which a given pixel is occupied by a front.
  }%
  \label{fig:frt-occurrence}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig05.pdf}
  \caption[Seasonal impact of fronts on Chl-\textit{a}]{
    Seasonal climatologies in the permanent subtropical biome (1st column), in the seasonal subtropical biome (2nd column) and in the subpolar biome (3rd column).
    (a-b-c)~\as{chl} median values (top row) over weak fronts (blue), strong fronts (green) and background (red). Gaps between the curves represent the local excess over fronts.
    (d-e-f)~Surface fraction occupied by fronts.
    (g-h-i)~Regional \as{chl} surplus at the scale of the biome.
    The surplus accounts for the local excess and for the number of fronts (see method).
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess in \as{chl} is larger over strong fronts than over weak fronts, but weak fronts are more numerous than strong fronts, resulting in a \as{chl} surplus that can be comparable or even larger for weak fronts.
  }%
  \label{fig:ts-climato}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig06.pdf}
  \caption[Local impact of front on Chl-\textit{a} in the permanent subtropical biome]{
    Permanent subtropical biome: local \as{chl} excess over fronts by range of latitudes.
    (a-c-e)~\as{chl} median values over weak fronts (blue) and background (red), (b-d-f)~corresponding local excess of \as{chl} in weak fronts computed as the relative difference of \as{chl} in fronts and in the background.
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess increases from south to north.
  }%
  \label{fig:latbands-s}
\end{figure*}

\begin{figure*}
  \includegraphics[width=12cm]{fig07.pdf}
  \caption[Local impact of front on Chl-\textit{a} in the seasonal subtropical biome]{
    Seasonal subtropical biome: local \as{chl} excess over fronts by range of latitudes.
    (a-c-e)~\as{chl} median values over weak fronts (blue), strong fronts (green) and background (red), (b-d-f)~corresponding local excess of \as{chl} in weak and strong fronts computed as the relative difference of \as{chl} in fronts and in the background.
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess is maximum at mid-latitudes.
  }%
  \label{fig:latbands-i}
\end{figure*}

\begin{figure*}
  \includegraphics[width=12cm]{fig08.pdf}
  \caption[Local impact of front on Chl-\textit{a} in the subpolar biome]{
    Subpolar biome: local \as{chl} excess over fronts by range of latitudes in the biome.
    (a-c-e)~\as{chl} median values over weak fronts (blue), strong fronts (green) and background (red), (b-d-f)~corresponding local excess of \as{chl} in weak and strong fronts computed as the relative difference of \as{chl} in fronts and in the background.
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess decreases from south to north.
  }%
  \label{fig:latbands-n}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{fig09.pdf}
  \caption[Recap of local and global impact of fronts on Chl-\textit{a}]{
    (a)~Annual mean local \as{chl} excess over fronts (in \%), sorted by latitudinal band (x"~axis), by biome (shape of symbol) and by front type (weak fronts in blue, strong fronts in green).
    (b)~Annual mean global surplus of \as{chl} (in \%) for each biome, sorted by front type.
  }%
  \label{fig:recap}
\end{figure*}

@@

\begin{figure}
  \includegraphics[width=8.3cm]{fig10.pdf}
  \caption[Comparison of bloom onset dates in the background and over fronts]{
    Subpolar biome. Comparison of bloom onset dates (in day of year) in the background (x"~axis) and over fronts (y"~axis), sorted by strength of fronts (shape of symbol), and latitudinal band (color).
    The line \(y=x\) is plotted in black.
    The distance between the black line and the dotted (respectively dashed) grey line is the measure of the average difference between weak (respectively strong) fronts and background.
    The bloom onset day propagates from south to north and starts earlier over fronts at all latitudes in the subpolar biome.
  }%
  \label{fig:bloom}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=12cm]{fig_suppl_01.pdf}
 \caption[Sensitivity of the seasonal impact of fronts on HI parameters]{
   Climatological mean of \as{chl} median values (top row) over weak fronts (blue), strong fronts (green) and background (red), surface fraction occupied by weak fronts and strong fronts (middle row), and global \as{chl} excess due to weak and strong fronts (bottom row).
   Each line represent a set of parameter with the bolder line indicating the retained set of parameters.
   The tested rolling window sizes are \qty{20}{\km}, \qty{30}{\km}, and \qty{40}{\km}.
   Different normalization coefficients are tested for a \qty{30}{\km} window size: double the variance, double the bimodality, and double the skewness.
 }%
 \label{fig:ts-sensitivity}
\end{figure*}

\begin{figure*}
  \includegraphics[width=12cm]{fig_suppl_02.pdf}
  \caption[Seasonal distributions of Chl-\textit{a}]{
    Distribution of \as{chl} of the year~2007 by seasons~(rows), for the three biomes~(columns), and for the background~(red), weak fronts~(blue) and strong fronts~(green).
    The median value of each distribution is indicated by a vertical line.
  }
  \label{fig:hist-chl}
\end{figure*}

>>>>

Our approach combines daily satellite SST data, which are used to detect fronts and sort them by their strength, with daily satellite surface \as{chl}, from which we derive anomalies over fronts.
For \as{chl}, we used the \abbrv{L3}~product distributed by ACRI"~ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (\as{seawifs}, \as{modis} Aqua \& Terra, \as{meris}, \as{viirs}-\as{snpp} \& \as{jpss}[\textlf{-1}], \as{olci}[\abbrv{-S3A}] \& \abbrv{S3B}) merged and reprocessed, available daily at \qty{4}{\km} resolution (\cite{garnesson_2019})\footnote{Dataset \dataname{chl_globcolour} décrit~\nref[section]{sec:donnees-chl_globcolour}, \nref{chp:methodes}}.

For SST, we used the \al{esa} \al{sst} \al{cci} analysis product version 2.1 (\cite{merchant_2019, good_2020})\footnote{Dataset \dataname{sst_esacci} décrit \nref[section]{sec:donnees-sst_esacci}, \nref{chp:methodes}}, also available daily at \qty{4}{\km} resolution over the period 2000--2020.
This product combines data from all available infrared sensors (\as{atsr}, \as{slstr}, and \as{avhrr}), ensuring good resolution where data are available, unlike other SST products which also include microwave and in-situ measurements, resulting in considerable smoothing of the SST field. %chktex 36
Where SST data is not available, spatial interpolation is performed to obtain a cloud free product which, at the cost of resolution on finer features, provides complete synoptic coverage of our large study area.
This interpolation tends to underestimate the detection of fronts, as the SST field is smoother over cloud-covered areas (\cite{merchant_2019}).
However, the combination of several sensors allows to reduce these areas to a minimum.
Furthermore, we have only considered cloud-free pixels for our analysis, which ensures that cloudy areas are not taken into account in our quantification.

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig01.pdf}
  \caption[Delimitation of biomes]{
    Delimitation of the three biomes in the open ocean Gulf Stream extension region: the Permanent Subtropical Biome (PSB, south of the dashed line at \latlon{32N}), the Seasonal Subtropical Biome (SSB, between \latlon{32N} and the meandering Gulf Stream northern wall on that day marked with the black contour), and the sub-Polar Biome (PB, north of the Gulf Stream northern wall).
    (a)~SST and (b)~\as{chl} snapshots on the 22 April 2007 (with data masked by clouds in white). The red line follows the \qty{1500}{\m} isobath.
    Data on the continental shelf (\qty{<1500}{\m}) is not considered here and have been masked.
    (c)~SST and (d)~\as{chl} distribution within each biome for the same day (PB:~blue, SSB:~yellow, PSB:~red).
    The black line in (c) shows the SST threshold value detected to delimit the Gulf Stream northern wall (see methods section).
  }%
  \label{fig:zone-separation}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig02.pdf}
  \caption[Example of front]{
    The SST, \as{chl}, and heterogeneity index (HI) of a front on the 7 July 2007.
    The plain and dashed contours correspond to HI values of 5 and 10.
    This front is categorized as weak.
    \as{chl} are elevated inside the front.
  }%
  \label{fig:zoom}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig03.pdf}
  \caption[Distribution of Chl-\textit{a} against HI]{
    Normalized distribution of the Heterogeneity Index (HI, red dashed line) within each biome, and distribution of \as{chl} as a function of HI (representing front strength), over the full period 2000--2020.
    Shown are the median value of the \as{chl} distributions (solid black line), and 1st and 3rd quartiles (dashed lines).
    Note that \pct{0.5} of pixels have outstanding large HI values and are not included here.
  }%
  \label{fig:chl-vs-hi}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig04.pdf}
  \caption[Occurrence map of strong and weak fronts]{
    Occurrence of (a) weak fronts and (b) strong fronts expressed as the percentage of time steps over the time series (2000--2020) for which a given pixel is occupied by a front.
  }%
  \label{fig:frt-occurrence}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig05.pdf}
  \caption[Seasonal impact of fronts on Chl-\textit{a}]{
    Seasonal climatologies in the permanent subtropical biome (1st column), in the seasonal subtropical biome (2nd column) and in the subpolar biome (3rd column).
    (a-b-c)~\as{chl} median values (top row) over weak fronts (blue), strong fronts (green) and background (red). Gaps between the curves represent the local excess over fronts.
    (d-e-f)~Surface fraction occupied by fronts.
    (g-h-i)~Regional \as{chl} surplus at the scale of the biome.
    The surplus accounts for the local excess and for the number of fronts (see method).
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess in \as{chl} is larger over strong fronts than over weak fronts, but weak fronts are more numerous than strong fronts, resulting in a \as{chl} surplus that can be comparable or even larger for weak fronts.
  }%
  \label{fig:ts-climato}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig06.pdf}
  \caption[Local impact of front on Chl-\textit{a} in the permanent subtropical biome]{
    Permanent subtropical biome: local \as{chl} excess over fronts by range of latitudes.
    (a-c-e)~\as{chl} median values over weak fronts (blue) and background (red), (b-d-f)~corresponding local excess of \as{chl} in weak fronts computed as the relative difference of \as{chl} in fronts and in the background.
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess increases from south to north.
  }%
  \label{fig:latbands-s}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig07.pdf}
  \caption[Local impact of front on Chl-\textit{a} in the seasonal subtropical biome]{
    Seasonal subtropical biome: local \as{chl} excess over fronts by range of latitudes.
    (a-c-e)~\as{chl} median values over weak fronts (blue), strong fronts (green) and background (red), (b-d-f)~corresponding local excess of \as{chl} in weak and strong fronts computed as the relative difference of \as{chl} in fronts and in the background.
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess is maximum at mid-latitudes.
  }%
  \label{fig:latbands-i}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig08.pdf}
  \caption[Local impact of front on Chl-\textit{a} in the subpolar biome]{
    Subpolar biome: local \as{chl} excess over fronts by range of latitudes in the biome.
    (a-c-e)~\as{chl} median values over weak fronts (blue), strong fronts (green) and background (red), (b-d-f)~corresponding local excess of \as{chl} in weak and strong fronts computed as the relative difference of \as{chl} in fronts and in the background.
    The plain lines represent the climatological mean, and the envelopes mark the standard deviation over the period 2000--2020.
    The excess decreases from south to north.
  }%
  \label{fig:latbands-n}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig09.pdf}
  \caption[Recap of local and global impact of fronts on Chl-\textit{a}]{
    (a)~Annual mean local \as{chl} excess over fronts (in \%), sorted by latitudinal band (x"~axis), by biome (shape of symbol) and by front type (weak fronts in blue, strong fronts in green).
    (b)~Annual mean global surplus of \as{chl} (in \%) for each biome, sorted by front type.
  }%
  \label{fig:recap}
\end{figure*}

@@

\begin{figure}
  \centering
  \includegraphics[width=8.3cm]{fig10.pdf}
  \caption[Comparison of bloom onset dates in the background and over fronts]{
    Subpolar biome. Comparison of bloom onset dates (in day of year) in the background (x"~axis) and over fronts (y"~axis), sorted by strength of fronts (shape of symbol), and latitudinal band (color).
    The line \(y=x\) is plotted in black.
    The distance between the black line and the dotted (respectively dashed) grey line is the measure of the average difference between weak (respectively strong) fronts and background.
    The bloom onset day propagates from south to north and starts earlier over fronts at all latitudes in the subpolar biome.
  }%
  \label{fig:bloom}
\end{figure}

@@

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig_suppl_01.pdf}
  \caption[Sensitivity of the seasonal impact of fronts on HI parameters]{
    Climatological mean of \as{chl} median values (top row) over weak fronts (blue), strong fronts (green) and background (red), surface fraction occupied by weak fronts and strong fronts (middle row), and global \as{chl} excess due to weak and strong fronts (bottom row).
    Each line represent a set of parameter with the bolder line indicating the retained set of parameters.
    The tested rolling window sizes are \qty{20}{\km}, \qty{30}{\km}, and \qty{40}{\km}.
    Different normalization coefficients are tested for a \qty{30}{\km} window size: double the variance, double the bimodality, and double the skewness.
  }%
 \label{fig:ts-sensitivity}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=12cm]{fig_suppl_02.pdf}
  \caption[Seasonal distributions of Chl-\textit{a}]{
    Distribution of \as{chl} of the year~2007 by seasons~(rows), for the three biomes~(columns), and for the background~(red), weak fronts~(blue) and strong fronts~(green).
    The median value of each distribution is indicated by a vertical line.
  }
  \label{fig:hist-chl}
\end{figure*}

### 2023-04-16 tex/intro.tex
Some fix

\begin{figure}[!h]
  \centering
  {%
    \setlength{\fboxsep}{0pt}%
    \framebox[\figwidth]{\insertfig{illustration/illus_preamble_smol.jpg}}
  }%
  \captionT{Les couleurs du Gulf Stream par satellite}{%
    \small
    Image en fausses couleurs de l'océan Atlantique Nord, prises par \as{modis} \sur~Terra le \frenchdate{2020}{02}{23}, inspiré par le travail de Norman Kuring pour \glshref{eo-illustration}.
  }
  \label{fig:oc-illustration}
\end{figure}

@@

Les effets de la submésoéchelle sur les cycles biogéochimiques présentés ci-dessus ne sont toutefois pas encore complétement élucidés, malgré les efforts des dernières décennies dans ce domaine.
En effet, l'étude des processus à ces échelles présente des difficultés.
Les diverses méthodologies usuelles peinent à donner une vue entière du problème.
Les structures qui nous intéressent sont fines et ont une courte période de vie, ce qui rend leur échantillonage (physique mais aussi biologique) un défi technique certain.
Capturer les variations rapide de la biologie en travers d'un front lors d'une campagne in-situ présente un défi technique certain (\cite{marrec_2018,tzortzis_2021}).
De plus, la courte période de vie des structures d'intérêt rend la tâche d'autant plus ardue.
Il en va de même pour l'imagerie satellite, pour laquelle la couverture nuageuse rend difficile de suivre temporellement une structure.
Par ailleurs, il est compliqué pour les satellites d'accéder à toute la biodiversité du phytoplancton, ainsi qu'au delà des premiers mètres en surface.
Enfin, les simulations numériques peuvent résoudre ces fines échelles, mais à un coût calculatoire élevé et prohibitif sur de trop longues durées (des projections climatiques \ab{par-ex}) (\cite{fox-kemper_2019}).

Malgré les limites évoquées, l'imagerie satellite donne l'opportunité d'observer ces effets à une large échelle (spatiale et temporelle), et permettrait d'en fournir une quantification.
Une telle démarche a été entreprise par \textcite{liu_2016}, dans la gyre subtropicale du Pacifique Nord.
Les autrices \encadra{en colocalisant les valeurs satellites de la \al{chl} avec les positions de fronts détectés à partir de la température de surface satellite} ont montré une augmentation des valeurs de \as{chl} dans les fronts par rapport au reste de la zone.
Nous étendons leur méthode, et l'appliquons à la région de l'Atlantique Nord, autour du Gulf Stream.

@@

Par ailleurs cette zone présente deux fronts permanents intenses, un front le long du talus continental, et le Gulf Stream, tous deux associés à des circulations verticales fortes et profondes \cite{flagg_2006,liao_2022}.
Mais l'on peut également trouver dans le reste de la zone des fronts éphémères en constante évolution (\cite{drushka_2019,sanchez-rios_2020}).

Afin de mieux définir nos objectifs, et avant de les présenter, nous dressons dans la section suivante un état de l'art des sujets concernés.
Nous exposerons ensuite le plan de ce manuscript.

@@

Commençons par définir notre objet d'étude principal: le phytoplancton.
Conforme à sa construction éthymologique, ce terme désigne un ensemble d'organismes pratiquant la photosynthèse (\emph{phyto-}) et dérivant au gré des courants (de \guil*{\hbox{\emph{πλαγκτόσ}}}: errer).
Comme pour les plantes terrestres, ces organismes sont influencés par leur environment.
Cependant cet environment est en perpétuel changement du fait des mouvements des masses d'eaux dans l'océan.
Dans cette section, on décrira d'abord succinctement les caractéristiques biologiques générales du phytoplancton et les moyens de l'observer.
Dans la section suivante, on s'intéressera plus en détail aux intéractions biophysiques entre le phytoplancton et les courants océaniques.

>>>>

\begin{figure}[!h]
  \centering
  {%
    \setlength{\fboxsep}{0pt}%
    \framebox[\figwidth]{\insertfig{illustration/illus_preamble_smol.jpg}}
  }%
  \captionT{Les couleurs du Gulf Stream par satellite}{%
    Image en fausses couleurs de l'océan Atlantique Nord, prises par \as{modis} \sur~Terra le \frenchdate{2020}{02}{23}, inspiré par le travail de Norman Kuring pour \glshref{eo-illustration}.
  }
  \label{fig:oc-illustration}
\end{figure}

@@

Les effets de la submésoéchelle sur les cycles biogéochimiques présentés ci-dessus ne sont toutefois pas encore complétement élucidés, malgré les efforts des dernières décennies dans ce domaine.
En effet, à ces échelles, les diverses méthodologies usuelles peinent à donner une vue entière du problème.
Les structures qui nous intéressent sont fines et ont une courte période de vie, ce qui rend leur échantillonage  un défi technique certain.
Les campagnes in-situ doivent déployer des moyens importants pour échantilloner les variables physiques et biologiques, ainsi que des techniques nouvelles (comme l'échantillonage lagrangien ou un cytomètre de flux automatisé \cite{thyssen_2015,marrec_2018,tzortzis_2021}).
Il en va de même pour l'imagerie satellite, pour laquelle la couverture nuageuse rend difficile de suivre temporellement une structure.
Par ailleurs, il est compliqué pour les satellites d'accéder à toute la biodiversité du phytoplancton, ainsi qu'au delà des premiers mètres en surface.
Enfin, les simulations numériques peuvent résoudre ces fines échelles, mais à un coût calculatoire élevé et prohibitif sur de trop longues durées (des projections climatiques \ab{par-ex}) (\cite{fox-kemper_2019}).

Malgré les limites évoquées, l'imagerie satellite donne l'opportunité d'observer ces effets de fine échelle, sur de larges régions et de longs temps, et permettrait ainsi de quantifier leurs effets globaux.
Une telle démarche a été entreprise par \textcite{liu_2016} dans la gyre subtropicale du Pacifique Nord.
Les autrices \encadra{en colocalisant les valeurs satellites de la \al{chl} avec les positions de fronts détectés à partir de la SST (température en surface)} ont montré une augmentation des valeurs de \as{chl} dans les fronts par rapport au reste de la zone.
Nous reprenons leur méthode de détection des fronts, consistant à calculer un indice d'hétérogénéité de la SST, et l'appliquons à la région de l'Atlantique Nord, autour du Gulf Stream.

@@

Par ailleurs cette zone présente deux fronts permanents intenses, un front le long du talus continental, et le Gulf Stream, tous deux associés à des circulations verticales fortes et profondes (\cite{flagg_2006,liao_2022}).
Mais l'on peut également trouver dans le reste de la zone des fronts moins intenses, d'une durée de vie de l'ordre de la semaine, et en constante évolution (\cite{drushka_2019,sanchez-rios_2020}).

Afin de mieux définir nos objectifs, et avant de les présenter, nous dressons dans la section suivante un état de l'art des sujets concernés.
Nous exposerons ensuite le plan de ce manuscrit.

@@

Commençons par définir notre objet d'étude principal: le phytoplancton.
Conforme à sa construction étymologique, ce terme désigne un ensemble d'organismes pratiquant la photosynthèse c'est-à-dire des plantes (\emph{phyto-}), et dérivant au gré des courants (de \emph{plankton}: errer).
Comme pour les plantes terrestres, ces organismes sont influencés par leur environment.
Cependant cet environment est en perpétuel changement du fait des mouvements des masses d'eaux dans l'océan.
Dans cette section, on décrira d'abord succinctement les caractéristiques biologiques générales du phytoplancton et les moyens de l'observer.
Dans la section suivante, on s'intéressera plus en détail aux intéractions biophysiques entre le phytoplancton et les courants océaniques.
